---
title: "1. Data Pre-Processing"
author: 'Team 1 (``Destroyers"): Jesus Vazquez, Leo Li, Ying Zhang, Tian Zhao'
date: "5/5/2021"
output: html_document
---

```{r setup, include=FALSE}
library(readr)
library(fastDummies)
library(mice)
library(caret)
library(dplyr)
```

First, we convert the categorical variables to dummy variable notations using reference cell coding.

```{r}
setwd("C:/Users/haoli/Desktop/735/Project")
train <- read.csv("aug_train.csv", na.strings = "")
train.work <- train[,-c(1,2)]
train.work$gender <- as.numeric(as.factor(train.work$gender))
train.work$relevent_experience <- as.numeric(as.factor(train.work$relevent_experience))
train.work$enrolled_university <- as.numeric(as.factor(train.work$enrolled_university))
train.work$education_level <- as.numeric(as.factor(train.work$education_level))
train.work$major_discipline <- as.numeric(as.factor(train.work$major_discipline))
train.work$experience <- as.numeric(as.factor(train.work$experience))
train.work$company_size <- as.numeric(as.factor(train.work$company_size))
train.work$company_type <- as.numeric(as.factor(train.work$company_type))
train.work$last_new_job <- as.numeric(as.factor(train.work$last_new_job))
train.work <- train.work[(train.work$education_level!=2)&(train.work$education_level!=5), ]
```

Second, we split the data set into a training and test set with a ratio of 70% to 30%.

```{r}
ind <- round(nrow(train.work)*0.7)
train.pre <- train.work[1:ind,]
test.pre <- train.work[(ind+1):nrow(train.work),]
```

Then, to prepare the data sets for complete-case analysis, we only keep the complete observations in the training and test data sets.

```{r}
# finalize test data
test.data <- test.pre[complete.cases(test.pre), ]
gender <- dummy_cols(test.data$gender)[, -c(1,2)]
colnames(gender) <- paste0('gender', colnames(gender))
relevent_experience <- dummy_cols(test.data$relevent_experience)[, -c(1,2)]
enrolled_university <- dummy_cols(test.data$enrolled_university)[, -c(1,2)]
colnames(enrolled_university) <- paste0('enrolled_university', colnames(enrolled_university))
education_level <- dummy_cols(test.data$education_level)[, -c(1,2)]
colnames(education_level) <- paste0('education_level', colnames(education_level))
major_discipline <- dummy_cols(test.data$major_discipline)[, -c(1,2)]
colnames(major_discipline) <- paste0('major_discipline', colnames(major_discipline))
experience <- dummy_cols(test.data$experience)[, -c(1,2)]
colnames(experience) <- paste0('experience', colnames(experience))
company_size <- dummy_cols(test.data$company_size)[, -c(1,2)]
colnames(company_size) <- paste0('company_size', colnames(company_size))
company_type <- dummy_cols(test.data$company_type)[, -c(1,2)]
colnames(company_type) <- paste0('company_type', colnames(company_type))
last_new_job <- dummy_cols(test.data$last_new_job)[, -c(1,2)]
colnames(last_new_job) <- paste0('last_new_job', colnames(last_new_job))
city_development_index <- test.data$city_development_index
training_hours <- test.data$training_hours
target <- test.data$target
test.data.final <- cbind(city_development_index, gender, relevent_experience, enrolled_university, education_level,
                         major_discipline, experience, company_size, company_type, last_new_job, training_hours, target)
write.csv(test.data.final, file = 'test.csv')

# cc training data
test.data <- train.pre[complete.cases(train.pre), ]
gender <- dummy_cols(test.data$gender)[, -c(1,2)]
colnames(gender) <- paste0('gender', colnames(gender))
relevent_experience <- dummy_cols(test.data$relevent_experience)[, -c(1,2)]
enrolled_university <- dummy_cols(test.data$enrolled_university)[, -c(1,2)]
colnames(enrolled_university) <- paste0('enrolled_university', colnames(enrolled_university))
education_level <- dummy_cols(test.data$education_level)[, -c(1,2)]
colnames(education_level) <- paste0('education_level', colnames(education_level))
major_discipline <- dummy_cols(test.data$major_discipline)[, -c(1,2)]
colnames(major_discipline) <- paste0('major_discipline', colnames(major_discipline))
experience <- dummy_cols(test.data$experience)[, -c(1,2)]
colnames(experience) <- paste0('experience', colnames(experience))
company_size <- dummy_cols(test.data$company_size)[, -c(1,2)]
colnames(company_size) <- paste0('company_size', colnames(company_size))
company_type <- dummy_cols(test.data$company_type)[, -c(1,2)]
colnames(company_type) <- paste0('company_type', colnames(company_type))
last_new_job <- dummy_cols(test.data$last_new_job)[, -c(1,2)]
colnames(last_new_job) <- paste0('last_new_job', colnames(last_new_job))
city_development_index <- test.data$city_development_index
training_hours <- test.data$training_hours
target <- test.data$target
test.data.final <- cbind(city_development_index, gender, relevent_experience, enrolled_university, education_level,
                         major_discipline, experience, company_size, company_type, last_new_job, training_hours, target)
write.csv(test.data.final, file = 'train_cc.csv')
```

The same test set will also be used for multiple imputation, and we prepare the training dataset for multiple imputation as follows.

```{R}
mi <- mice(train.pre)
for (i in 1:5){
  mi1 <- complete(mi, action = i, include = F)
  test.data <- mi1[complete.cases(mi1), ]
  gender <- dummy_cols(test.data$gender)[, -c(1,2)]
  colnames(gender) <- paste0('gender', colnames(gender))
  relevent_experience <- dummy_cols(test.data$relevent_experience)[, -c(1,2)]
  enrolled_university <- dummy_cols(test.data$enrolled_university)[, -c(1,2)]
  colnames(enrolled_university) <- paste0('enrolled_university', colnames(enrolled_university))
  education_level <- dummy_cols(test.data$education_level)[, -c(1,2)]
  colnames(education_level) <- paste0('education_level', colnames(education_level))
  major_discipline <- dummy_cols(test.data$major_discipline)[, -c(1,2)]
  colnames(major_discipline) <- paste0('major_discipline', colnames(major_discipline))
  experience <- dummy_cols(test.data$experience)[, -c(1,2)]
  colnames(experience) <- paste0('experience', colnames(experience))
  company_size <- dummy_cols(test.data$company_size)[, -c(1,2)]
  colnames(company_size) <- paste0('company_size', colnames(company_size))
  company_type <- dummy_cols(test.data$company_type)[, -c(1,2)]
  colnames(company_type) <- paste0('company_type', colnames(company_type))
  last_new_job <- dummy_cols(test.data$last_new_job)[, -c(1,2)]
  colnames(last_new_job) <- paste0('last_new_job', colnames(last_new_job))
  city_development_index <- test.data$city_development_index
  training_hours <- test.data$training_hours
  target <- test.data$target
  test.data.final <- cbind(city_development_index, gender, relevent_experience, enrolled_university, education_level,
                           major_discipline, experience, company_size, company_type, last_new_job, training_hours, target)
  write.csv(test.data.final, file = paste0('train_mi_', i, '.csv'))
}
```

Finally, the down sampling is conducted as follows.

```{r}
# downsample
dat <- read.csv("train_cc.csv")
set.seed(1)
down_dat <- downSample(x = dat %>% select(-c("target")),
                       y = factor(dat$target)) %>%  rename(target=Class)
write.csv(down_dat, file = 'train_cc_downsample.csv')
```
